Profitability Prediction of Consulting Projects in the Construction Industry

### 2016-08-24 finalwriteup/draft.rmd
merge with home sick work

1. Storing a collection of projects with key variable values
2. Once a new project or case arrives, similar cases are retrieved. This can be achieved either algorithmically with a distance function applied to the variables, or manually by users reviewing past cases. The nearest neighbour algorithm has been used to find the distance between cases in previous studies. It is programmed to calculate the Euclidean distance in n-dimensional space between cases, where each variable is a dimension [@Kumar2007].
3. The cost of the new project is estimated by extrapolating characteristics of the similar cases to the new case. An experienced decision maker can also do this either algorithmically or manually.

@@

Linear Regression is one of the simplest and most popular statistical prediction methods. An equation for a line is learned, which is defined by a constant and coefficients multiplied by each explanatory variable. This can be mathematically represented by the equation below:

@@

[@Hastie2009]

@@

The most popular method to find the coefficients is to minimise the residual sum of squares (ordinary least squared method). The residuals represent the Y distance of each point from the line (predicted values), and the residual sum of squares is the sum of the squared value of each point's residual. It can be represented by the equation below:

@@

If structured correctly, Linear Regression results can show the importance of the explanatory variables. The variables must be first standardised to one another (i.e. scaled to the same mean and standard deviation). Then, their coefficients can be directly compared [@Putler2012]. The magnitude of the coefficients (the slopes) indicate how *much* a variable changes with the response variable. However, if two or more predictor variables are correlated, the values of their coefficients are unstable. More specifically, this means that if the a model has two correlated variables, and is re-run with a slightly different set of data, the coefficients for the two variables may be quite different. This is because explanatory variables are assumed to be independent, and if variables are correlated, it breaks this assumption. On the other hand, correlated variables do not impact the predictive accuracy of the model [@Putler2012].

@@

Nevertheless, the theory behind Linear Regression can be adapted to the prediction of probabilities pertaining to a binary response variable (represented by 1 or 0). First, a sigmoid function is assumed as the functional form of the probabilities against an example variable X. This is because it is pragmatic and works well: sigmoid functions can be bound by 0 and 1, and a logit transformation of the probability results in a linear relationship between the log odds and X. Refer to the equation below where *p* represents probability:

@@

[@Macdonald1975]

Because of the linear relationship, an equation can be fit to the log odds of the binary response variable against each of the explanatory variables using linear regression. Then, the linear equation describing the log odds can be transformed back to probability by taking the inverse log, i.e. the exponential and rearranging:

@@

[@Macdonald1975]

@@

Decision trees have a different theoretical foundation to traditional prediction methods such as regression. They are created by a progression of binary splits of the data set that attempt to group similar response variable values. Before a split is made, each explanatory variable and each value within the explanatory variables is tested for how 'purely' it partitions the response variable into two subsets. For example, the explanatory variable 'gender' might split a response variable 'hair length' of a class of students into two subsets of students with more similar hair lengths. This similarity is called 'purity' and can be measured using several formulas. An example for regression (a problem with a numeric response variable) is:

@@

![Simple 4 node Bayesian Network](images/lit_review/BN_1.png)

@@

![4 node Bayesian Network with conditional independencies](images/lit_review/BN_2.png)

@@

In this way, Bayesian Networks reduce computations required to find the probability of a unique combination given other explanatory variable values while still taking into account many conditional dependencies that Naive Bayes, for example, cannot [@Barber2012]. Also, some equations for a combination are now clones of another combination and do not require recalculation. Calculating the conditional probability tables for each variable is still much more computationally expensive than Naive Bayes (as Naive Bayes is the simplest form of a Bayesian Network) and variables with many categories dramatically increases the expense [@Zhang2004].

@@

The method for finding the boundary, however, is different to Logistic Regression. The aim is to find the hyperplane with the widest clear margin between data points from the two categories, while allowing a certain number of misclassifications (this is usually optimised for the problem). To do this, the equation for the hyperplane is expressed as follows:

@@

The plane is expressed with vectors because they can be defined such that $\boldsymbol{w}$ is perpendicular to $\boldsymbol{x}$. This aids in optimisation calculations to find the hyperplane with the widest margin, i.e. two hyperplanes with the largest perpendicular distance between them that contains less than the allowable number of misclassified points [@Fletcher2009]. The plot below demonstrates this concept and has no misclassifications:

@@

The purpose of permuting is to mimic the absence of that variable. Now the accuracy of the trees where the covariate has been permuted can be compared against the original Random Forest. For a categorical response variable, refer to the equation describing variable importance score below:

@@

RMSE_b = baseline root mean squared error
y = predicted 'return per dollar'
y = mean 'return per dollar' for al projects

@@

$$
\begin{aligned}
RMSE_b = \textrm{baseline root mean squared error} \\
n &= \textrm{number of data points} \\
\widehat{y} &= \textrm{predicted values of the response variable} \\
\bar y &= \textrm{mean 'return per dollar' for all projects} \\
\end{aligned}
$$

@@

```{r, fig.height = 4.3, echo = FALSE, fig.cap = c("Distribution of the difference in RMSE between randomForest predictions (built on 6 core variables) and a baseline predictor that uses the mean 'return per dollar' as its prediction. If the Random Forest model was effective, the difference would be greater than 0.")}

@@

```{r, fig.height = 4.3, echo = FALSE, fig.cap = c("Distribution of the difference in RMSE between randomForest predictions (built on all variables) and a baseline predictor that uses the mean 'return per dollar' as its prediction. If the Random Forest model is effective, the difference should be greater than 0.")}

@@

The p-value for these models and the aforementioned null hypothesis is `r t.test(results.rf2$diff, alternative = "greater", mu = 0)$p.value %>% format(digits = 2)`. The null hypothesis was rejected which contrasted the matching ANOVA test. On the right hand tail of the graph above, the RMSE is improved by almost \$0.20 but also worsens by over \$0.10 per dollar on the left tail. A density plot of the unaltered RMSE values for the models are shown below:

```{r, fig.height = 4.3, echo = FALSE, fig.cap = c("Distribution of RMSE's of 'return per dollar' from 100 randomForest models built on random subsamples of the variables")}

@@

Both ANOVA and Random Forest models produced a similar lowest RMSE value, but performed differently when directly comparing models built on the same list of variables. Both methods predicted at best within an average of \$0.48 for the response variable, 'return per dollar', using either core variables or a sample of core and incomplete variables. Random Forest predictions were much more accurate than ANOVA's with a mix of core (almost complete) and less variables while Random Forest and ANOVA performed similarly with the core variables only. There was much less data available for training the models that used less-complete variables so it was interesting that Random Forest performed considerably better (RMSE of \$0.50 for Random Forest vs. \$2 for ANOVA). The main difference between ensemble tree methods and ANOVA is that ANOVA assumes linear relationships between variables (some of which are normalised) whereas trees are capable of capturing non-linear relationships. It is possible the less complete variables had non-linear relationships with the response variable. Random Forest is also an ensemble method while ANOVA is not. Each Random Forest is the average of 500 single trees, whereas each ANOVA model is a single model only. 50 Random Forest models produce improved overall performance.

@@

```{r reduced_violin, fig.height = 5.5, fig.cap = c("Violin plot vertically illustrating the distribution of AUC values from each of the methods when predicting profit/loss. Subsets of the data were used for Logistic Regression and randomForests in order to provide datasets without missing values.")}

@@

This chapter first presents the full range of method results in terms of improvements to the case study's bottom line. A business decision making scenario was created so that profit curves based on the decision rule could be built and analysed. From the profit curves, optimal probability thresholds could be derived for each method (individual and blended methods). These threshold points allowed the probability output from each model to be converted to categorical output (profit or loss) which in turn could generate confusion matrix statistics for each method. The confusion matrix statistics are discussed after the profit curve results and discussion. Then, output from the method which produced the best profit curve was used to create two subsets of profitable and loss mkaing projects. The characteristics of these subsets are compared and contrasted to gauge any overall trends. Finally, alternate ways of framing the business decision scenario are proposed.

@@

```{r profit_curve, fig.height = 6, fig.cap = c("Profit Curves summarising results from 100 models of 9 methods: 3 simple blends, 3 complex blends, and the original 3 best methods.")}

@@

ggplot(finance.plot, aes(x= threshold, y= profit.ratio, colour = method)) + 
        # geom_point(alpha = 0.005, shape = 16, size = 1) + 
        geom_hline(yintercept = 100, colour = "darkgray", linetype = 2) + 
        geom_hline(yintercept = 0, colour = "darkgray", linetype = 1) +
        stat_summary(fun.y = mean, 
                     fun.ymin = function(x) mean(x) - 1.96*sd(x),
                     fun.ymax = function(x) mean(x) + 1.96*sd(x),
                     fill = "gray", geom= "ribbon", alpha = 0.7, colour = NA) +
        stat_summary(fun.y = mean, geom="line") +
        scale_y_continuous(breaks = seq(0,110, by = 20)) + 
        scale_x_continuous(breaks = seq(0,1, by = .2)) +  
        theme(legend.position = "none",
              panel.background = element_rect(fill = 'gray95')) +
        labs(x = "Probability Threshold", 
             y = "Ratio of Model Profit to Historical Profit (%)",
             title = "Profit Curve for Blended and Individual Models") + 
        facet_wrap(~method, 
                   labeller = labeller( method = c("average" = "Simple Average Blend",
                                       "orig.boost"= "Orig. Boosted Trees",
                                       "orig.log" = "Orig. Logistic Regression",
                                       "orig.rf" = "Orig. Random Forest",
                                       "simp.log" = "Simple Log. Reg. Blend",
                                       "simp.boost" = "Simple Boosted Tree Blend",
                                       "comp.boost"= "Complex Boosted Tree Blend",
                                       "comp.rf"= "Complex randomForest Blend",
                                       "FWLS" = "FWLS"
                                       )))

@@

```{r av_profit_curve, fig.height = 4.5, fig.cap = c("Profit Curve of the best performing method: the simple Logistic Regression Blended Method")}

@@

```{r}
# Results - over 5-pages, up to 20 pages
# 
# Guidelines:
# * Outlines what you found out in relation to your research questions or hypotheses, presented in figures and in written text.
# * Results contain the facts of your research. Often you will include a brief comment on the significance of key results, with the expectation that more generalised comments about results will be made in the Discussion section.
# 
# Which results to discuss
# * restate research questions and hypotheses. can we predict profitability? ofocurse you can make a prediction, 
# real question might be is it worthwhile for the business to engage in this activity? we will look at this as a bottom line analysis. acknowledge that it may be worthwhile for other reasons such as confidence and transparency in historic projects.
# 
# * bottom line analysis - run 1000 times to get confidence interval around the line.
#     * should the business adopt the model?
#     
# * to address the problem of trust and create insight to the past data - nearest neighbour algorithm. This also mimics the CBR approach
```

@@

To review method performance further, confusion matrix statistics were scrutinised. For each method, 100 full data sets of results were generated via 5-fold cross validation. Each data set of probability results differed to the next because of the random division into the five-fold data sets used for training and testing. For this analysis, nine full data sets of results was chosen that produced the mean profitability ratio at each of the nine method's optimal threshold points. These were deemed a 'typical' set of predictions for each method. Then, the confusion matrix for each set of results at their respective thresholds could be calculated. The plot below shows the breakdown of FN's and TP's on true loss making projects. If the project was predicted correctly as loss making, this was a TP, and the job was rejected (in the fabricated decision scenario). If it was predicted incorrectly as a profit-making job, this was a false negative (FN), and the project was completed, resulting in a loss for the business.

@@

All below plots and comments summarise results from one Simple Blended Logistic Regression Model. The results from this model were chosen because it represented the mean improvement in profits output by this blending method. The plot below illustrates how the ratio of profitable to unprofitable jobs compare in the full project population vs. the rejected projects.

@@

Industry's lack of uptake of cost estimation models was marked as a significant issue in the literature [@Akintoye2000]. Decision makers did not sufficiently trust the model output which were not directly connected to their experiences. In many studies, CBR had more traction than more accurate models because decision makers could justify the output. Clearly this is an important aspect of the study to address and could be managed at the point where the decision maker engages with the algorithm: the user interface. This interface should do two things well:

@@

Two strategies were adopted to facilitate user trust in the model's output: illustration of the model structure and a nearest neighbour graph. Algorithms were originally selected that could reveal their predictive structure by either variable importance and/or variable relationships. The first tab in the application displays several variable dependency plots extracted from a typical Boosted Tree model. Understanding the plot would require some training. The y-axis represents the probability a project will be loss making after taking into account average effects from variables other than the x-axis variable.

@@

![Boosted Tree Partial dependency plot for Timespan embedded in the user interface](images/app/timespan_dep.png)

Secondly, the nearest neighbour algorithm is automatically run with the predictive model and its results are displayed in the second main tab. The aim was to provide the user with a cross check of the algorithm's results by finding projects with the most similar features to the one currently being assessed. This triggers memories from the user's past experiences on similar projects and promotes communication with other managers who delivered the 'nearest neighbours'. Some screen shots of the nearest neigbhour graphic are presented below.

@@

![Tables below the chart in the application give specific details of each of the nearest neighbours](images/app/NN_table_top.png)

![The nearest neighbour details are split into sections such as 'Finances' and 'Staff Details'](images/app/NN_table.png)

The similar projects are calculated using a different method from the blended algorithm but it is important to have a form of cross-checking and engage the user's own experiences. Further development and testing is required to assess how intuitive the results are for users to interpret, how much users trust the algorithm output given the variable plots and nearest neighbour charts, and finally how trustworthiness may be further improved.

@@

First, successfully estimating project variables is critical in translating this research into tangible business benefits. The data represents the company's records of *completed* projects so initially expected variable values were not documented. For example, if a project was originally expected to last 1.5 years but instead took 2 years, the company data only retains the two year time span. This prompts the question, how well could a user predict the variables before beginning a project? Uncertainty has been managed by binning variables such as 'total amount invoiced' and 'time span' to 5 or 6 broad categories (i.e. "1 day - 3 weeks", "3 weeks - 2.5 months", etc). However, two remaining categories: team size and percent of project completed by professional level employees were left as continuous variables. It would be interesting to test how well business managers could guess the aforementioned four variables at the start of a project.

@@

The general aim was to use statistical techniques to model the profitability of projects for a case study consulting business using their internal CRM data. This was rigorously completed by approaching the prediction of 'profitability' in two ways: as a regression problem predicting the degree of profitability and as a binary classification problem predicting either profit or loss. Several statistical and machine learning approaches were applied to both problems including Naive Bayes, Bayesian Networks, Linear Regression, Random Forests, gradient Boosted Trees as well as multiple methods of blending the individual models.

@@

Lastly, hypothesis 2 proposed that the best predictive model developed from hypothesis 1 could be shown to improve the overall profitability (bottom line) of the case study business. This theory was tested on a range of predictive models because AUC scores were not directly linked to profitability once the models were applied to a business decision scenario. The models output a probability, so an optimal probability threshold needed to be found that would dictate whether a job was classified 'profitable' or 'loss making'. A range of threshold values were trialed, and a decision framework was created where if a project was scored below the threshold it was accepted while projects above the threshold were rejected. If a project was accepted, the profits and losses were summed while if a project was rejected the profits and losses were forfeited. A perfect model would reject all loss making jobs and accept all profitable jobs, thereby improving overall profits. Final results showed the simple Logistic Regression blend of the individual Logistic Regression, Random Forest and Boosted Tree models improved profits the most. The 95% confidence interval for these improvements was between `r best.fin %>% filter(method == "simp.log") %>% .$lower_bound_95CI - 100`% and `r best.fin %>% filter(method == "simp.log") %>% .$upper_bound_95CI - 100`% using a probability threshold of `r best.t` (approximately `r ((simp.l %>% filter(mean.iter>best.t) %>% nrow)/nrow(simp.l)*100) %>% round(1)`% of projects).

These results contribute significantly to the research in cost estimation in three ways: the applied methods, the decision framework, and the appeal to user trust. Ensemble tree methods and blending had been applied minimally to cost estimation to date. Past research tested regression models and to some extent Neural Networks and CBR. The advantage of ensemble tree methods is that they provide insight into the model's structure while predicting at a similar level to Neural Networks. Next, a clear framework for how a business could tangibly improve profits was developed and proven successful. Previous studies have verified predictive accuracy but stopped short of how the algorithm would affect decisions and what the measured benefits would be. Lastly, the decision framework was progressed even further into a prototype application where user trust was addressed. Previous studies had shown that despite decades of promising research in cost estimation, decision maker's still mainly use an 'inside view' during fee estimation. This was addressed by providing intuitive plots of the model structure and a parallel nearest neighbour graph of similar internal projects to provide a clear and intuitive 'outside view'. Further work is required to test user confidence in the output.

Another topic identified for future research was to test how well managers estimate project input variables. Estimation of the time span, team size, total invoiced amount, and percentage of hours completed by professionals in particular should be tested. Initial estimates of input variables could also be kept in the database for future analysis.

Overall, this work has successfully built a mathematical blend of Logistic Regression, Random Forest, and Boosted Tree models, built on a consulting company's internal project data. This blended model can predict whether a project will be profitable or not and in a reasonable business decision framework, can guide managers in rejecting financially risky projects and improving final profitability.

>>>>

1. Storing a collection of projects with key variable values
2. Once a new project or case arrives, similar cases are retrieved. This can be achieved either algorithmically with a distance function applied to the variables, or manually by users reviewing past cases. The Nearest Neighbour algorithm has been used to find the distance between cases in previous studies. It is programmed to calculate the Euclidean distance in n-dimensional space between cases, where each variable is a dimension [@Kumar2007].
3. The cost of the new project is estimated by extrapolating characteristics of the similar cases to the new case. An experienced decision maker can also do this either algorithmically or manually.

@@

Linear Regression is one of the simplest and most popular statistical prediction methods. An equation for a line is learned, which is defined by a constant and coefficients multiplied by each explanatory variable. This can be mathematically represented by the equation below [@Hastie2009]:

@@

@@

The most popular method to find the coefficients is to minimise the residual sum of squares (ordinary least squared method). The residuals represent the Y distance of each point from the line (predicted values), and the residual sum of squares is the sum of the squared value of each point's residual. It can be represented by the equation below [@Hastie2009]:

@@

If structured correctly, Linear Regression results can show the importance of the explanatory variables. The variables must be first standardised to one another (i.e. scaled to the same mean and standard deviation). Then, their coefficients can be directly compared [@Putler2012]. The magnitude of the coefficients (the slopes) indicates how *much* a variable changes with the response variable. However, if two or more predictor variables are correlated, the values of their coefficients are unstable. More specifically, this means that if a model has two correlated variables, and is re-run with a slightly different set of data, the coefficients for the two variables may be quite different. This is because explanatory variables are assumed to be independent, and if variables are correlated, it breaks this assumption. On the other hand, correlated variables do not impact the predictive accuracy of the model [@Putler2012].

@@

Nevertheless, the theory behind Linear Regression can be adapted to the prediction of probabilities pertaining to a binary response variable (represented by 1 or 0). First, a sigmoid function is assumed as the functional form of the probabilities against an example variable X. This is because it is pragmatic and works well: sigmoid functions can be bound by 0 and 1, and a logit transformation of the probability results in a linear relationship between the log odds and X. Refer to the equation below where *p* represents probability [@Macdonald1975]:

@@

Because of the linear relationship, an equation can be fit to the log odds of the binary response variable against each of the explanatory variables using linear regression. Then, the linear equation describing the log odds can be transformed back to probability by taking the inverse log, i.e. the exponential and rearranging [@Macdonald1975]:

@@

@@

Decision trees have a different theoretical foundation to traditional prediction methods such as regression. They are created by a progression of binary splits of the data set that attempt to group similar response variable values. Before a split is made, each explanatory variable and each value within the explanatory variables is tested for how 'purely' it partitions the response variable into two subsets. For example, the explanatory variable 'gender' might split a response variable 'hair length' of a class of students into two subsets of students with more similar hair lengths. This similarity is called 'purity' and can be measured using several formulas. An example for regression (a problem with a numeric response variable) is [@Hastie2009]:

@@

![Simple 4-Node Bayesian Network](images/lit_review/BN_1.png)

@@

![4-Node Bayesian Network with conditional independencies](images/lit_review/BN_2.png)

@@

In this way, Bayesian Networks reduce computations required to find the probability of a unique combination given other explanatory variable values while still taking into account many conditional dependencies that Naive Bayes, for example, cannot [@Barber2012]. Also, some equations for a combination are now clones of another combination and do not require recalculation. Calculating the conditional probability tables for each variable is still much more computationally expensive than Naive Bayes (as Naive Bayes is the simplest form of a Bayesian Network) and variables with many categories dramatically increase the expense [@Zhang2004].

@@

The method for finding the boundary, however, is different to Logistic Regression. The aim is to find the hyperplane with the widest clear margin between data points from the two categories, while allowing a certain number of misclassifications (this is usually optimised for the problem). To do this, the equation for the hyperplane is expressed as follows [@Fletcher2009]:

@@

The plane is expressed with vectors because they can be defined such that $\boldsymbol{w}$ is perpendicular to $\boldsymbol{x}$. This aids optimisation calculations to find the hyperplane with the widest margin, i.e. two hyperplanes with the largest perpendicular distance between them that contains less than the allowable number of misclassified points [@Fletcher2009]. The plot below demonstrates this concept and has no misclassifications:

@@

The purpose of permuting is to mimic the absence of that variable. Now the accuracy of the trees where the covariate has been permuted can be compared against the original Random Forest. For a categorical response variable, refer to the equation describing variable importance score below [@Breiman2005]:

@@

@@

$$
\begin{aligned}
RMSE_b &= \textrm{baseline root mean squared error} \\
n &= \textrm{number of data points} \\
\widehat{y} &= \textrm{predicted values of the response variable} \\
\bar y &= \textrm{mean 'return per dollar' for all projects} \\
\end{aligned}
$$

@@

```{r, fig.height = 4.3, echo = FALSE, fig.cap = c("Distribution of the difference in RMSE between Random Forest predictions (built on 6 core variables) and a baseline predictor that uses the mean 'return per dollar' as its prediction. If the Random Forest model was effective, the difference would be greater than 0.")}

@@

```{r, fig.height = 4.3, echo = FALSE, fig.cap = c("Distribution of the difference in RMSE between Random Forest predictions (built on all variables) and a baseline predictor that uses the mean 'return per dollar' as its prediction. If the Random Forest model is effective, the difference should be greater than 0.")}

@@

The p-value for these models and the aforementioned null hypothesis is `r t.test(results.rf2$diff, alternative = "greater", mu = 0)$p.value %>% format(digits = 2)`. The null hypothesis was rejected which contrasted the matching ANOVA test. On the right hand tail of the graph above, the RMSE is improved by almost \$0.20 but also worsens by over \$0.10 per dollar on the left tail. A density plot of the unaltered RMSE values for the models is shown below:

```{r, fig.height = 4.3, echo = FALSE, fig.cap = c("Distribution of RMSE's of 'return per dollar' from 100 Random Forest models built on random subsamples of the variables")}

@@

Both ANOVA and Random Forest models produced a similar lowest RMSE value, but performed differently when directly comparing models built on the same list of variables. Both methods predicted at best within an average of \$0.48 for the response variable, 'return per dollar', using either core variables or a sample of core and incomplete variables. Random Forest predictions were much more accurate than ANOVA's with a mix of core (almost complete) and less complete variables while Random Forest and ANOVA performed similarly with the core variables only. There was much less data available for training the models that used less-complete variables so it was interesting that Random Forest performed considerably better (RMSE of \$0.50 for Random Forest vs. \$2 for ANOVA). The main difference between ensemble tree methods and ANOVA is that ANOVA assumes linear relationships between variables (some of which are normalised) whereas trees are capable of capturing non-linear relationships. It is possible the less complete variables had non-linear relationships with the response variable. Random Forest is also an ensemble method while ANOVA is not. Each Random Forest is the average of 500 single trees, whereas each ANOVA model is a single model only. 50 Random Forest models produce improved overall performance.

@@

```{r reduced_violin, fig.height = 5.5, fig.cap = c("Violin plot vertically illustrating the distribution of AUC values from each of the methods when predicting profit/loss. Subsets of the data were used for Logistic Regression and Random Forests in order to provide datasets without missing values.")}

@@

This chapter first presents the full range of method results in terms of improvements to the case study's bottom line. A business decision making scenario was created so that profit curves based on the decision rule could be built and analysed. From the profit curves, optimal probability thresholds could be derived for each method (individual and blended methods). These threshold points allowed the probability output from each model to be converted to categorical output (profit or loss) which in turn could generate confusion matrix statistics for each method. The confusion matrix statistics are discussed after the profit curve results and discussion. Then, output from the method which produced the best profit curve was used to create two subsets of profitable and loss making projects. The characteristics of these subsets are compared and contrasted to gauge any overall trends. Finally, alternate ways of framing the business decision scenario are proposed.

@@

```{r profit_curve, fig.height = 6, fig.cap = c("Profit curves summarising results from 100 models of 9 methods: 3 simple blends, 3 complex blends, and the original 3 best methods.")}

@@

ggplot(finance.plot, aes(x= threshold, y= profit.ratio, colour = method)) + 
        # geom_point(alpha = 0.005, shape = 16, size = 1) + 
        geom_hline(yintercept = 100, colour = "darkgray", linetype = 2) + 
        geom_hline(yintercept = 0, colour = "darkgray", linetype = 1) +
        stat_summary(fun.y = mean, 
                     fun.ymin = function(x) mean(x) - 1.96*sd(x),
                     fun.ymax = function(x) mean(x) + 1.96*sd(x),
                     fill = "gray", geom= "ribbon", alpha = 0.7, colour = NA) +
        stat_summary(fun.y = mean, geom="line") +
        scale_y_continuous(breaks = seq(0,110, by = 20)) + 
        scale_x_continuous(breaks = seq(0,1, by = .2)) +  
        theme(legend.position = "none",
              panel.background = element_rect(fill = 'gray95')) +
        labs(x = "Probability Threshold", 
             y = "Ratio of Model Profit to Historical Profit (%)",
             title = "Profit Curve for Blended and Individual Models") + 
        facet_wrap(~method, 
                   labeller = labeller( method = c("average" = "Simple Average Blend",
                                       "orig.boost"= "Orig. Boosted Trees",
                                       "orig.log" = "Orig. Logistic Regression",
                                       "orig.rf" = "Orig. Random Forest",
                                       "simp.log" = "Simple Log. Reg. Blend",
                                       "simp.boost" = "Simple Boosted Tree Blend",
                                       "comp.boost"= "Complex Boosted Tree Blend",
                                       "comp.rf"= "Complex Random Forest Blend",
                                       "FWLS" = "FWLS"
                                       )))

@@

```{r av_profit_curve, fig.height = 4.5, fig.cap = c("Profit curve of the best performing method: the simple Logistic Regression Blended Method")}

@@

```{r}
# Results - over 5-pages, up to 20 pages
# 
# Guidelines:
# * Outlines what you found out in relation to your research questions or hypotheses, presented in figures and in written text.
# * Results contain the facts of your research. Often you will include a brief comment on the significance of key results, with the expectation that more generalised comments about results will be made in the Discussion section.
# 
# Which results to discuss
# * restate research questions and hypotheses. can we predict profitability? ofocurse you can make a prediction, 
# real question might be is it worthwhile for the business to engage in this activity? we will look at this as a bottom line analysis. acknowledge that it may be worthwhile for other reasons such as confidence and transparency in historic projects.
# 
# * bottom line analysis - run 1000 times to get confidence interval around the line.
#     * should the business adopt the model?
#     
# * to address the problem of trust and create insight to the past data - Nearest Neighbour algorithm. This also mimics the CBR approach
```

@@

To review method performance further, confusion matrix statistics were scrutinised. For each method, 100 full data sets of results were generated via 5-fold cross validation. Each data set of probability results differed to the next because of the random division into the five-fold data sets used for training and testing. For this analysis, nine full data sets of results were chosen that produced the mean profitability ratio at each of the nine method's optimal threshold points. These were deemed a 'typical' set of predictions for each method. Then, the confusion matrix for each set of results at their respective thresholds could be calculated. The plot below shows the breakdown of FN's and TP's on true loss making projects. If the project was predicted correctly as loss making, this was a TP, and the job was rejected (in the fabricated decision scenario). If it was predicted incorrectly as a profit-making job, this was a false negative (FN), and the project was completed, resulting in a loss for the business.

@@

All below plots and comments summarise results from one Simple Blended Logistic Regression Model. The results from this model were chosen because it represented the mean improvement in profits output by this blending method. The plot below illustrates how the ratio of profitable to unprofitable jobs compares in the full project population vs. the rejected projects.

@@

Industry's lack of uptake of cost estimation models was marked as a significant issue in the literature [@Akintoye2000]. Decision makers did not sufficiently trust the model output which was not directly connected to their experiences. In many studies, CBR had more traction than more accurate models because decision makers could justify the output. Clearly this is an important aspect of the study to address and could be managed at the point where the decision maker engages with the algorithm: the user interface. This interface should do two things well:

@@

Two strategies were adopted to facilitate user trust in the model's output: illustration of the model structure and a Nearest Neighbour graph. Algorithms were originally selected that could reveal their predictive structure by either variable importance and/or variable relationships. The first tab in the application displays several variable dependency plots extracted from a typical Boosted Tree model. Understanding the plot would require some training. The y-axis represents the probability a project will be loss making after taking into account average effects from variables other than the x-axis variable.

@@

![Boosted Tree partial dependency plot for 'Timespan' embedded in the user interface](images/app/timespan_dep.png)

Secondly, the Nearest Neighbour algorithm is automatically run with the predictive model and its results are displayed in the second main tab. The aim was to provide the user with a cross check of the algorithm's results by finding projects with the most similar features to the one currently being assessed. This triggers memories from the user's past experiences on similar projects and promotes communication with other managers who delivered the 'Nearest Neighbours'. Some screen shots of the nearest neigbhour graphic are presented below.

@@

![Tables below the chart in the application give specific details of each of the Nearest Neighbours](images/app/NN_table_top.png)

![The Nearest Neighbour details are split into sections such as 'Finances' and 'Staff Details'](images/app/NN_table.png)

The similar projects are calculated using a different method from the blended algorithm but it is important to have a form of cross-checking and to engage the user's own experiences. Further development and testing is required to assess how intuitive the results are for users to interpret, how much users trust the algorithm output given the variable plots and Nearest Neighbour charts, and finally how trustworthiness may be further improved.

@@

First, successfully estimating project variables is critical in translating this research into tangible business benefits. The data represents the company's records of *completed* projects so initially expected variable values were not documented. For example, if a project was originally expected to last 1.5 years but instead took 2 years, the company data only retains the two year time span. This prompts the question, how well could a user predict the variables before beginning a project? Uncertainty has been managed by binning variables such as 'total amount invoiced' and 'time span' to 5 or 6 broad categories (i.e. "1 day - 3 weeks", "3 weeks - 2.5 months", etc.). However, two remaining categories: team size and percent of project completed by professional level employees were left as continuous variables. It would be interesting to test how well business managers could guess the aforementioned four variables at the start of a project.

@@

The general aim was to use statistical techniques to model the profitability of projects for a case study consulting business using their internal CRM data. This was rigorously completed by approaching the prediction of 'profitability' in two ways: as a regression problem predicting the degree of profitability and as a binary classification problem predicting either profit or loss. Several statistical and machine learning approaches were applied to both problems including Naive Bayes, Bayesian Networks, Linear Regression, Random Forests, gradient Boosted Trees as well as multiple methods of blending the individual models' output.

@@

Lastly, hypothesis 2 proposed that the best predictive model developed from hypothesis 1 could be shown to improve the overall profitability (bottom line) of the case study business. This theory was tested using a range of predictive models because AUC scores were not directly linked to profitability once the models were applied to a business decision scenario. The models output probability scores, so an optimal probability threshold needed to be found that would dictate whether a job was classified as 'profitable' or 'loss making'. A range of threshold values were trialed, and a decision framework was created where if a project was scored below the threshold it was accepted while projects above the threshold were rejected. If a project was accepted, the profits and losses were summed while if a project was rejected the profits and losses were forfeited. A perfect model would reject all loss making jobs and accept all profitable jobs, thereby improving overall profits. Final results showed the simple Logistic Regression blend of the individual Logistic Regression, Random Forest and Boosted Tree models improved profits the most. The 95% confidence interval for these improvements was between `r best.fin %>% filter(method == "simp.log") %>% .$lower_bound_95CI - 100`% and `r best.fin %>% filter(method == "simp.log") %>% .$upper_bound_95CI - 100`% using a probability threshold of `r best.t` (approximately `r ((simp.l %>% filter(mean.iter>best.t) %>% nrow)/nrow(simp.l)*100) %>% round(1)`% of projects).

These results contribute significantly to the research in cost estimation in three ways: the applied methods, the decision framework, and the appeal to user trust. Ensemble tree methods and blending had been applied minimally to cost estimation previously, and past research had favoured Linear Regression, Neural Networks and CBR. Ensemble tree methods are advantageous because they provide insight into the model's structure while predicting at a similar level to Neural Networks. Next, a clear framework for how a business could tangibly improve profits was developed and proven successful. Previous studies have verified predictive accuracy but stopped short of how the algorithm would affect decisions and what the measured benefits would be. Lastly, the decision framework was progressed even further into a prototype application where user trust was addressed. Past surveys have reported that despite decades of promising research in cost estimation, decision maker's still mainly use an 'inside view' during fee estimation. This was addressed by displaying intuitive plots of the model structure as well as a Nearest Neighbour graph of similar internal projects that provide an 'outside view'. Further work is required to test user confidence in the output.

Another topic identified for future research was to test how well managers estimate some numeric project input variables. In particular, time span, team size, total invoiced amount, and percentage of hours completed by professionals should be tested as these variables were calculated post project completion. Time span and total invoiced amount were discretised into wide categories, which should be easier for a manager to choose between. Initial estimates of input variables could also be kept in the database for future analysis.

Overall, this work has successfully built a mathematical blend of Logistic Regression, Random Forests, and Boosted Tree models, built on a consulting company's internal project data. This blended model can predict whether a project will be profitable or not and in a reasonable decision framework, can guide managers in rejecting financially risky projects and improving profitability of the business.
